{\fontfamily{ptm}\selectfont A Parameter Free Accelerated Proximal Gradient Method Without Restarting }

### 2025-02-25 Notebook LaTEX/R-WAPG/paper3.tex
>w<, fix cringe shit

\begin{abstract} 
    \noindent
    We study the accelerated proximal gradient method where the sequence that controls the momentum term doesn't follow the Nesterov's rule. 
    The paper proposed R-WAPG, a generic algorithm that unifies the convergence results for strongly convex and convex where the extrapolation constant is characterized by a sequence that is much weaker than the Nesterov's rule. 
    The R-WAPG provides a unified framework for several notable Euclidean variants of FISTA and verify their convergence. 
    In addition, we also study the convergence rate of strongly convex objective with a constant momentum term. 
    We reformulate R-WAPG so that it doesn't require any parameter without using the idea of restarting which we call ``Free R-WAPG". 
    Explorative numerical experiments were conducted to show its competitive convergence. 
\end{abstract}

@@

        We see that it admits quadratic growth over the minimizer $Ty$, therefore for all $x \in \RR^n$ it follows that: 

@@

            \widetilde{\mathcal M}^{L^{-1}}(Ty; y)

@@

            \frac{L}{2}\Vert x - Ty\Vert^2

@@

            \mathcal M^{L^{-1}}(Ty; y) 

@@

            \frac{L}{2}\Vert x - Ty\Vert^2
            + D_f(Ty; y)

@@

            - \frac{L}{2}\Vert x - Ty\Vert^2
            + D_f(Ty; y)
                F(x) - F(Ty) 

@@

                \frac{L}{2}\Vert Ty - y\Vert^2

@@

            + D_f(Ty; y)
            - \frac{L}{2}\Vert x - Ty\Vert^2
                F(x) - F(Ty) 

@@

                    \Vert x - Ty + Ty - y\Vert^2

@@

                    \Vert y - Ty\Vert^2

@@

                    \Vert x - Ty\Vert^2 + 
                    2\langle x - Ty, Ty - y\rangle

@@

            + D_f(Ty; y)
            - \frac{L}{2}\Vert x - Ty\Vert^2

@@

                F(x) - F(Ty) + \frac{L}{2}\Vert x - Ty\Vert^2 
                - L\langle  x - Ty, y - Ty\rangle

@@

            + D_f(Ty; y)
            - \frac{L}{2}\Vert x - Ty\Vert^2
            F(x) - F(Ty)
            - \langle L(y - Ty), x - Ty\rangle

@@

            + D_f(Ty, y).

@@

            F(x) - F(Ty)
            - \langle L(y - Ty), x - Ty\rangle

@@

            + D_f(Ty, y)

@@

            + \frac{L}{2}\Vert y - Ty\Vert^2
            F(x) - F(Ty)
            - \langle L(y - Ty), x - y  + y - Ty\rangle

@@

            + \frac{L}{2}\Vert y - Ty\Vert^2
            F(x) - F(Ty) - \langle L(y - Ty), x - y\rangle - \frac{\mu}{2}\Vert x - y \Vert^2
            - \frac{L}{2}\Vert y - Ty\Vert^2. 

@@

\section{Stepwise formulation of weak accelerated proximal gradient}\label{sec:stepwise-stuff}
    The goal of this section is to build the R-WAPG algorithm which is described in the Definition \ref{def:wapg} of the next section. 
    \par 
    Definition \ref{def:stepwise-wapg} describes one iteration of the R-WAPG algorithm. 
    It defines the procedure of generating $x_{k + 1}, v_{k + 1}$ given any $x_k, v_k$. 
    Proposition \ref{prop:stepwise-lyapunov} states the inequality \eqref{ineq:stepwise-lya-2}  of a decreasing quantity involving $F(x_k), F(x_{k + 1})$ at each iteration.
    \par
    The assumption below is crucial to describing the convergence rate of the R-WAPG and it requires the convexity assumption on $F$. 
    \begin{assumption}
        Given $x_k, y_k, v_k$ where $k \in \mathbb Z_+$, we define the following quantities
        \begin{align}
            g_k &\defeq L(y_k - T_L(y_k)), 
            \label{eqn:grad-map}
            \\
            l_F(x; y_k) &\defeq F(T_L(y_k)) + \langle g_k, x - y_k\rangle + \frac{1}{2L}\Vert g_k\Vert^2, 
            \label{eqn:lower-linearization}
            \\
            \epsilon_{k} &\defeq F(x_k) - l_F(x_k; y_k), 
            \label{eqn:regret}
        \end{align}
        % To see $\epsilon_k \ge 0$, use Theorem \ref{thm:prox-grad-ineq} and let $y = y_k, x = x_k$, by the definition $g_k = L(y_k - T_L(y_k))$ it has: 
        % \begin{align*}
        %     F(x_k) - F(T_L(y_k))
        %     - \langle L(y_k - T_L(y_k)),x_k - y_k \rangle
        %     - \frac{L}{2}\Vert y_k - T_Ly_k\Vert^2
        %     - \frac{\mu}{2}\Vert x_k - y_k\Vert^2
        %     &\ge 0
        %     \\
        %     \iff 
        %     F(x_k) - F(T_L(y_k))
        %     - \langle g_k,x_k - y_k \rangle
        %     - \frac{1}{2L}\Vert g_k\Vert^2
        %     &\ge 0. 
        % \end{align*}
    \end{assumption}

@@

            g_k &= \mathcal G_L y_k, \label{eqn:stepwise-wapg-eqn3}

@@

            x_{k + 1} &= T_L y_k. \label{eqn:stepwise-wapg-eqn5}

@@

        With $\epsilon_k$ given by \eqref{eqn:regret}, define: 

@@

        Let $\epsilon_k$ be given by \eqref{eqn:regret}. 

@@

                    y_k - L^{-1} \mathcal G_L y_k, 

@@

                    \right) - \frac{1}{L\alpha_{k}}\mathcal G_L y_k. 

@@

                    & x_{k + 1} \defeq y_k - L^{-1}\mathcal G_Ly_k, 

@@

                - \alpha_k\hat \gamma_{k + 1}^{-1}\mathcal G_L y_k

@@

                - \frac{1}{L\alpha_{k}}\mathcal G_L y_k. 

@@

                    x_{k + 1} &= y_k - L^{-1}\mathcal G_Ly_k, 

@@

    \begin{theorem}[Fixed momentum APG]\label{thm:fixed-momentum-fista}
        Assume $L > \mu > 0$, let a pair of constant R-WAPG sequence: $(\alpha_k)_{k \ge0}, (\rho_k)_{k \ge 0}$ be given by Lemma \ref{lemma:constant-rwapg-seq}.
        Define $q := \mu/L$ and for any fixed $r \in \left(\sqrt{q}, \sqrt{q^{-1}}\right)$, let $\alpha_k := \alpha = r \sqrt{q}$ be the constant R-WAPG sequence. 
        Consider the algorithm with a constant momentum specified by the following: 
        \begin{tcolorbox}
            Define $\theta = \left(1 - r^{-1}\sqrt{q}\right)(1 - r\sqrt{q})(1 - q)^{-1}$. 
            \\
            Initialize $y_1 = x_1$; for $k = 1, 2, \ldots, N$, update: 
            \begin{align*}
                &x_{k + 1} = y_k + L^{-1}\mathcal G_L y_k, 
                \\
                & y_{k + 1} = x_{k + 1} + \theta(x_{k + 1} - x_k). 
            \end{align*}
        \end{tcolorbox}
        Then the algorithm generates $(x_k)_{k \ge 1}$ such that $F(x) - F(x^*)$ convergences at a rate of $\mathcal O\left(\left(1 - \min\left(\mu/(\alpha L), \alpha\right)\right)^k\right)$. 
    \end{theorem}
    \begin{proof}
        The constant sequence $(\alpha_k)_{k \ge 0}, (\rho_k)_{k \ge 0}$ is a valid R-WAPG sequence by Lemma \ref{lemma:constant-rwapg-seq}. 
        The algorithm presented here is the same as the R-WAPG momentum form as given by Definition \ref{def:r-wapg-momentum-form} because the term $\theta$ from Lemma \ref{lemma:constant-rwapg-seq} matches up with Definition \ref{def:r-wapg-momentum-form} by (ii) in Lemma \ref{lemma:constant-rwapg-seq}. 
        Then, Proposition \ref{prop:wapg-convergence} applies,and it follows that: 
        \begin{align*}
            & F(x_{k + 1}) - F^* + \frac{L\alpha_k^2}{2}\Vert v_{k + 1} - x^*\Vert^2
            \\
            &= 
            F(x_{k + 1}) - F^* + \frac{L\alpha_k^2}{2}\Vert x_{k + 1}  - x^* + (\alpha^{-1} - 1)(x_{k + 1} - x_k)\Vert^2
            \\
            &\le 
            \left(
                \prod_{i = 0}^{k - 1} \max(1, \rho_{k})
            \right)
            \left(
                \prod_{i = 1}^{k} \left(1  - \alpha_i\right)
            \right)
            \left(
                F(x_1) - F(x^*) + \frac{L\alpha_0^2}{2}\Vert v_1 - x^*\Vert^2
            \right)
            \\
            &= \left(
                \prod_{i = 1}^{k - 1} \max(1, \rho)(1 - \alpha)
            \right)
            \left(
                F(x_1) - F(x^*) + \frac{L\alpha_0^2}{2}\Vert v_1 - x^*\Vert^2
            \right)
            \\
            &= \left(1 - \min\left(\frac{\mu}{\alpha L}, \alpha\right)\right)^k
            \left(
                F(x_1) - F(x^*) + \frac{L\alpha_0^2}{2}\Vert x_1 - x^*\Vert^2
            \right). 
            & \text{Lemma \ref{lemma:constant-rwapg-seq} (v)}
        \end{align*}
        Going from the second last to the last equality, we used the initialization condition $y_1 = x_1$ specified in Proposition \ref{prop:r-wapg-momentum-repr}, therefore it has $v_1 = x_1$. 
    \end{proof}
    \begin{remark}
        When $r = 1$, the algorithm described above is exactly the same as the V-FISTA algorithm specified in (10.7.7) of Beck's book \cite{beck_first-order_2017}. 
        The above theorem gives convergence rate for a much relax choice of momentum parameter $\theta$ generalizing the convergence of constant step size FISTA to all convex stronlgy convex function and all possible $\alpha_k \in (\mu/L, 1)$. 
        In addition, the convergence claim of R-WAPG holds for any $\mu$ smaller than the strong convexity modulus of $f$ that is not zero making the bounds $r \in (\sqrt{q}, \sqrt{q^{-1}})$ looser. 
    \end{remark}

@@

\section{Discussions}
    This paper provides the results of an upper bound on the optimality gap $F(x_k) - F^*$ for smooth plus nonsmooth composite optimization problem with convex objectives where under the much weaker assumption for the momentum parameter specified by Definition \ref{def:rwapg-seq}. 
    The proposed R-WAPG unifies the convergence proof for major Euclidean variants of FISTA with strong convexity parameter $\mu \ge 0$. 
    To the best of our knowledge, there is no framework had been identified yet in the literatures that proves the convergences of several variants (with and without strong convexity parameter) with momentum sequences that doesn't follow the Nesterov's rule. 
    \par
    In addition, we proposed explorative numerical experiments for Free R-WAPG (Algorithm \ref{alg:free-rwapg}) showcasing a parameter free algorithm that doesn't use the idea of restarting to achieve competitive convergence rate. 
    Unfortunately a precise description of the convergence rate of Free R-WAPG remains a mystery. 
    Our convergence framework gives an upper bound using the sequence $(\alpha_k)_{k \ge 0} \in (\mu/L, 1)$, but it requires knowing what the sequence would be in advanced given the initial conditions and the objective function. 
    \par
    From a practical perspective, Algorithm \ref{alg:free-rwapg} shows competitive results of convergence of strongly convex function.  
    The algorithm behaves in way having characteristics of both V-FISTA and M-FISTA which is fascinating. 
    In addition, of being parameter free, the algorithm works well with auto-differentiation framework because it requires the function value at the same point where the gradient is evaluated at $y_k$. 
    We do note that in the case when function is not strongly convex, the function value converges without $\Vert \mathcal G_L(y_k)\Vert$ converging.

>>>>

\begin{abstract} 
    \noindent
    We study the accelerated proximal gradient method where the sequence that controls the momentum term doesn't follow the Nesterov's rule. 
    The paper proposes Relaxed Weak Accelerated Proximal Gradient(R-WAPG), a generic algorithm that unifies the convergence results for strongly convex and convex problems where the extrapolation constant is characterized by a sequence that is much weaker than the Nesterov's rule. 
    The R-WAPG provides a unified framework for several notable Euclidean variants of FISTA and verify their convergence. 
    In addition, we also provide the convergence rate of strongly convex objective with a constant momentum term. 
    We reformulate R-WAPG so that it doesn't require any parameter, without using the idea of restarting, which we call ``Free R-WAPG". 
    Explorative numerical experiments were conducted to show its competitive convergence. 
\end{abstract}

@@

        We see that it admits quadratic growth over the minimizer $\bar y = T(y)$, therefore for all $x \in \RR^n$ it follows that: 

@@

            \widetilde{\mathcal M}^{L^{-1}}(\bar y; y)

@@

            \frac{L}{2}\Vert x - \bar y\Vert^2

@@

            \mathcal M^{L^{-1}}(\bar y; y) 

@@

            \frac{L}{2}\Vert x - \bar y\Vert^2
            + D_f(\bar y; y)

@@

            - \frac{L}{2}\Vert x - \bar y\Vert^2
            + D_f(\bar y; y)
                F(x) - F(\bar y) 

@@

                \frac{L}{2}\Vert \bar y - y\Vert^2

@@

            + D_f(\bar y; y)
            - \frac{L}{2}\Vert x - \bar y\Vert^2
                F(x) - F(\bar y) 

@@

                    \Vert x - \bar y + \bar y - y\Vert^2

@@

                    \Vert y - \bar y\Vert^2

@@

                    \Vert x - \bar y\Vert^2 + 
                    2\langle x - \bar y, \bar y - y\rangle

@@

            + D_f(\bar y; y)
            - \frac{L}{2}\Vert x - \bar y\Vert^2

@@

                F(x) - F(\bar y) + \frac{L}{2}\Vert x - \bar y\Vert^2 
                - L\langle  x - \bar y, y - \bar y\rangle

@@

            + D_f(\bar y; y)
            - \frac{L}{2}\Vert x - \bar y\Vert^2
            F(x) - F(\bar y)
            - \langle L(y - \bar y), x - \bar y\rangle

@@

            + D_f(\bar y, y).

@@

            F(x) - F(\bar y)
            - \langle L(y - \bar y), x - \bar y\rangle

@@

            + D_f(\bar y, y)

@@

            + \frac{L}{2}\Vert y - \bar y\Vert^2
            F(x) - F(\bar y)
            - \langle L(y - \bar y), x - y  + y - \bar y\rangle

@@

            + \frac{L}{2}\Vert y - \bar y\Vert^2
            F(x) - F(\bar y) - \langle L(y - \bar y), x - y\rangle - \frac{\mu}{2}\Vert x - y \Vert^2
            - \frac{L}{2}\Vert y - \bar y\Vert^2. 

@@

\section{Stepwise formulation of weak accelerated proximal gradient}\label{sec:stepwise-stuff}
    The goal of this section is to build the R-WAPG algorithm which is described in the Definition \ref{def:wapg} of the next section. 
    \par 
    Definition \ref{def:stepwise-wapg} describes one iteration of the R-WAPG algorithm. 
    It defines the procedure of generating $x_{k + 1}, v_{k + 1}$ given any $x_k, v_k$. 
    Proposition \ref{prop:stepwise-lyapunov} states the inequality \eqref{ineq:stepwise-lya-2}  of a decreasing quantity involving $F(x_k), F(x_{k + 1})$ at each iteration.
    \par
    The assumption below is crucial to describing the convergence rate of the R-WAPG and it requires the convexity assumption on $F$. 
    \begin{assumption}\label{ass:regret-gap}
        Given $x_k, y_k, v_k$ where $k \in \mathbb Z_+$, we define the following quantities
        \begin{align}
            g_k &\defeq L(y_k - T_L(y_k)), 
            \label{eqn:grad-map}
            \\
            l_F(x; y_k) &\defeq F(T_L(y_k)) + \langle g_k, x - y_k\rangle + \frac{1}{2L}\Vert g_k\Vert^2, 
            \label{eqn:lower-linearization}
            \\
            \epsilon_{k} &\defeq F(x_k) - l_F(x_k; y_k), 
            \label{eqn:regret}
        \end{align}
        % To see $\epsilon_k \ge 0$, use Theorem \ref{thm:prox-grad-ineq} and let $y = y_k, x = x_k$, by the definition $g_k = L(y_k - T_L(y_k))$ it has: 
        % \begin{align*}
        %     F(x_k) - F(T_L(y_k))
        %     - \langle L(y_k - T_L(y_k)),x_k - y_k \rangle
        %     - \frac{L}{2}\Vert y_k - T_Ly_k\Vert^2
        %     - \frac{\mu}{2}\Vert x_k - y_k\Vert^2
        %     &\ge 0
        %     \\
        %     \iff 
        %     F(x_k) - F(T_L(y_k))
        %     - \langle g_k,x_k - y_k \rangle
        %     - \frac{1}{2L}\Vert g_k\Vert^2
        %     &\ge 0. 
        % \end{align*}
    \end{assumption}

@@

            g_k &= \mathcal G_L (y_k), \label{eqn:stepwise-wapg-eqn3}

@@

            x_{k + 1} &= T_L(y_k). \label{eqn:stepwise-wapg-eqn5}

@@

        With $\epsilon_k, y_k$ given by Assumption \ref{ass:regret-gap}, define: 

@@

        Let $\epsilon_k, y_k$ be given by Assumption \ref{ass:regret-gap}. 

@@

                    y_k - L^{-1} \mathcal G_L (y_k), 

@@

                    \right) - \frac{1}{L\alpha_{k}}\mathcal G_L (y_k). 

@@

                    & x_{k + 1} \defeq y_k - L^{-1}\mathcal G_L (y_k), 

@@

                - \alpha_k\hat \gamma_{k + 1}^{-1}\mathcal G_L (y_k)

@@

                - \frac{1}{L\alpha_{k}}\mathcal G_L (y_k). 

@@

                    x_{k + 1} &= y_k - L^{-1}\mathcal G_L (y_k), 

@@

    \begin{theorem}[Fixed momentum APG]\label{thm:fixed-momentum-fista}
        Assume $L > \mu > 0$, let a pair of constant R-WAPG sequence: $(\alpha_k)_{k \ge0}, (\rho_k)_{k \ge 0}$ be given by Lemma \ref{lemma:constant-rwapg-seq}.
        Define $q := \mu/L$ and for any fixed $r \in \left(\sqrt{q}, \sqrt{q^{-1}}\right)$, let $\alpha_k := \alpha = r \sqrt{q}$ be the constant R-WAPG sequence. 
        Consider the algorithm with a constant momentum specified by the following: 
        \begin{tcolorbox}
            Define $\theta = \left(1 - r^{-1}\sqrt{q}\right)(1 - r\sqrt{q})(1 - q)^{-1}$. 
            \\
            Initialize $y_1 = x_1$; for $k = 1, 2, \ldots, N$, update: 
            \begin{align*}
                &x_{k + 1} = y_k + L^{-1}\mathcal G_L (y_k)
                , 
                \\
                & y_{k + 1} = x_{k + 1} + \theta(x_{k + 1} - x_k). 
            \end{align*}
        \end{tcolorbox}
        Then the algorithm generates $(x_k)_{k \ge 1}$ such that $F(x) - F(x^*)$ convergences at a rate of $\mathcal O\left(\left(1 - \min\left(\mu/(\alpha L), \alpha\right)\right)^k\right)$. 
    \end{theorem}
    \begin{proof}
        The constant sequence $(\alpha_k)_{k \ge 0}, (\rho_k)_{k \ge 0}$ is a valid R-WAPG sequence by Lemma \ref{lemma:constant-rwapg-seq}. 
        The algorithm presented here is the same as the R-WAPG momentum form as given by Definition \ref{def:r-wapg-momentum-form} because the term $\theta$ from Lemma \ref{lemma:constant-rwapg-seq} matches up with Definition \ref{def:r-wapg-momentum-form} by (ii) in Lemma \ref{lemma:constant-rwapg-seq}. 
        Then, Proposition \ref{prop:wapg-convergence} applies,and it follows that: 
        \begin{align*}
            & F(x_{k + 1}) - F^* + \frac{L\alpha_k^2}{2}\Vert v_{k + 1} - x^*\Vert^2
            \\
            &= 
            F(x_{k + 1}) - F^* + \frac{L\alpha_k^2}{2}\Vert x_{k + 1}  - x^* + (\alpha^{-1} - 1)(x_{k + 1} - x_k)\Vert^2
            \\
            &\le 
            \left(
                \prod_{i = 0}^{k - 1} \max(1, \rho_{k})
            \right)
            \left(
                \prod_{i = 1}^{k} \left(1  - \alpha_i\right)
            \right)
            \left(
                F(x_1) - F(x^*) + \frac{L\alpha_0^2}{2}\Vert v_1 - x^*\Vert^2
            \right)
            \\
            &= \left(
                \prod_{i = 1}^{k - 1} \max(1, \rho)(1 - \alpha)
            \right)
            \left(
                F(x_1) - F(x^*) + \frac{L\alpha_0^2}{2}\Vert v_1 - x^*\Vert^2
            \right)
            \\
            &= \left(1 - \min\left(\frac{\mu}{\alpha L}, \alpha\right)\right)^k
            \left(
                F(x_1) - F(x^*) + \frac{L\alpha_0^2}{2}\Vert x_1 - x^*\Vert^2
            \right). 
            & \text{Lemma \ref{lemma:constant-rwapg-seq} (v)}
        \end{align*}
        Going from the second last to the last equality, we used the initialization condition $y_1 = x_1$ specified in Proposition \ref{prop:r-wapg-momentum-repr}, therefore it has $v_1 = x_1$. 
    \end{proof}
    \begin{remark}
        When $r = 1$, the algorithm described above is exactly the same as the V-FISTA algorithm specified in (10.7.7) of Beck's book \cite{beck_first-order_2017}. 
        The above theorem gives convergence rate for a much relax choice of momentum parameter $\theta$ generalizing the convergence of constant step size FISTA to all convex stronlgy convex function and all possible $\alpha_k \in (\mu/L, 1)$. 
        In addition, the convergence claim of R-WAPG holds for any $\mu$ smaller than the strong convexity modulus of $f$ that is not zero making the bounds $r \in (\sqrt{q}, \sqrt{q^{-1}})$ looser. 
    \end{remark}

@@

\section{Discussions}
    This paper provides the results of an upper bound on the optimality gap $F(x_k) - F^*$ for smooth plus nonsmooth composite optimization problem with convex objectives where under much weaker assumption for the momentum parameter. 
    The proposed R-WAPG unifies the convergence proof for major Euclidean variants of FISTA with strong convexity parameter $\mu \ge 0$. 
    To the best of our knowledge, there is no framework had been identified yet in the literatures that proves the convergences of several variants (with and without strong convexity parameter) with momentum sequences that doesn't follow the Nesterov's rule. 
    \par
    In addition, we proposed explorative numerical experiments for Free R-WAPG (Algorithm \ref{alg:free-rwapg}) showcasing a parameter free algorithm that doesn't use the idea of restarting to achieve competitive convergence rate. 
    Unfortunately a precise description of the convergence rate of Free R-WAPG remains a mystery. 
    It will be interesting to explore this line of research. 
    Our convergence framework gives an upper bound using the sequence $(\alpha_k)_{k \ge 0} \in (\mu/L, 1)$, but it requires knowing what the sequence would be in advanced given the initial conditions and the objective function. 
    \par
    From a practical perspective, Algorithm \ref{alg:free-rwapg} shows competitive results of convergence for strongly convex function.  
    The algorithm behaves in way having characteristics of both V-FISTA and M-FISTA which is fascinating. 
    In addition, of being parameter free, the algorithm works well with auto-differentiation framework because it requires the function value at the same point where the gradient is evaluated at $y_k$. 
    We do note that in the case when function is not strongly convex, the function value converges without $\Vert \mathcal G_L(y_k)\Vert$ converging. 
    This will be explored more in future work.
